Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex
minor-changes

{\color  {Myblue}Section~\ref{sec:defs}  introduces   the  syntax  
  of the  programming  language  considered  in our  analysis and describes describes  a small-step  operational semnatics for it.   We  describe our  liveness
  analysis  for  this  lazy  language  in  Section~\ref{sec:liveness}.
  Section~\ref{sec:computing}  describes   the  formulation   of  the
  liveness as a grammar and its approximation to DFA. It also contains
  a proof  showing the undecidability  of the precise liveness  in our
  formulation which necessitates the approximation to DFA.   Section~\ref{sec:GC-scheme} discusses  different ways
  of  handling  closures and  their  pros  and  cons.  We  report  our
  experimental  results  in Section~\ref{sec:experiments}  along  with
  some observations.  Section~\ref{sec:relatedwork} discusses previous
  work   related    to   garbage    collection   and    liveness   and
  Section~\ref{sec:conclusion}   discusses  possible   extensions  and
  concludes the paper.  }
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of  closures through  the $\LET$  construct.  $\LET$s  in our
language  are lazy;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
\IN\,\, e$, $x$ may occur in  $s$. This enables creation of graph-like
structures in  a pure language. The  restriction of \LET\ to  a single
definition  is for  ease  of  exposition---generalization to  multiple
definitions does not add conceptual difficulties.  We further restrict
each variable in a program to  be distinct, so that no scope shadowing
occurs---this  simplifies  the  reasoning   about  the  program.   The
$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
evaluation of closures.

@@

\begin{figure}[t]
\footnotesize
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\, e_\mainpgm
    \hspace{5em} \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \mbox{\em --- function definition} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{2em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) & \\
       (\CDR\,\, x) & \mbox{\em --- selectors} \\
       (\NULLQ\,\, x) & \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester
and generic arithmetic} \\
       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
  \caption{The syntax of our language}\label{fig:lang-syntax}
\figrule
\normalsize
\end{figure}

@@

The closest  to our approach  is the liveness based  garbage collector
implemented in~\cite{asati14lgc}. We extend  their work to handle lazy
evaluation and closures.

>>>>

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language  considered  in our  analysis  and  a small-step  operational
semnatics  for it.   We present  our liveness  analysis for  this lazy
language  in Section~\ref{sec:liveness}.   Section~\ref{sec:computing}
describes  the  formulation of  the  liveness  as  a grammar  and  its
approximation  to   DFA.  It  also   contains  a  proof   showing  the
undecidability  of  the  precise  liveness in  our  formulation  which
necessitates  the approximation  to DFA.   Section~\ref{sec:GC-scheme}
discusses different ways of handling closures and their pros and cons.
We  report our  experimental results  in Section~\ref{sec:experiments}
along with some observations.  Section~\ref{sec:relatedwork} discusses
previous  work   related  to  garbage  collection   and  liveness  and
Section~\ref{sec:conclusion}   discusses   possible   extensions   and
concludes the paper.
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of  closures through  the $\LET$  construct.  $\LET$s  in our
language  are lazy;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
\IN\,\, e$, $x$ may occur in  $s$. This enables creation of graph-like
structures in  a pure language. The  restriction of \LET\ to  a single
definition  is for  ease  of  exposition---generalization to  multiple
definitions does not add conceptual difficulties.  We further restrict
each variable in a program to  be distinct, so that no scope shadowing
occurs---this  simplifies  the  reasoning   about  the  program.   The
$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
evaluation of closures.

@@

\begin{figure}[t]
\renewcommand{\arraystretch}{1.5}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\, e_\mainpgm
    \hspace{5em} \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \mbox{\em --- function definition} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{2em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) & \\
       (\CDR\,\, x) & \mbox{\em --- selectors} \\
       (\NULLQ\,\, x) & \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester
and generic arithmetic} \\
       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
  \caption{The syntax of our language}\label{fig:lang-syntax}
\figrule
\normalsize
\end{figure}

@@

Another  interesting  research  in  this   field  is  the  concept  of
Simplifiers~\cite{ONeill}. They are  described as lightweight daeomons
which improve the efficiency of the program in general and the garbage
collector in particular. Most of the simplifications mentioned in this
work are subsumed by a liveness  based collector. We feel this work is
orthogonal and can augment our  liveness based collector.  The closest
to our  approach is the  liveness based garbage  collector implemented
in~\cite{asati14lgc}. We  extend their work to  handle lazy evaluation
and closures.
